Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2019-01-25 BMC-Response-Round1.tex
sync

In particular, the reviews helped us to state the goal of this work more clearly and improve its presentation. While we demonstrate the applicability of our approach to negative RNA design, which  aims at high and similar probability of the target structures (here measured as 'MultiDefect'), the focus of this work is the positive design of sequences that fold into multiple target structures with highly specific properties. We discuss this positive design, in particular, for specific folding energies (for each single target structure) and GC content. Moreover, as we like to emphasize, the method is even more general, which we see as a very important strength of our approach; it is prepared to handle many additional constraints and sequence features, using the same mechanism, stemming from practical use cases.

Naturally, we hope that this method turns out to be highly useful for single- and multi-target negative design (and various other applications, like generating background models for assessing significance) in future applications, but---while we present preliminary results for negative design (based on generating seed sequences for the MODENA benchmark)---a more in-depth evaluation of its use in negative design is beyond the scope of this paper. This seems even more acceptable, since the principal benefits of controlled sampling  have been demonstrated by RNABlueprint. We apologize for potential misunderstandings about the objectives of the manuscript and hope to better clarify these aspects in our revision.

@@

\Answer{
  We thank the reviewer for voicing her concerns, which motivated us to work on clarifying the practical use of the method/tool RNARedPrint in several ways; we have carefully edited the entire manuscript having these issues in mind.
  In the methods section, we tried to improve the general readability especially of our overview of the method and central problem.
  Moreover, we structured and rephrased the Results section more clearly and adapted our contributions outline and conclusion accordingly. In Results, we introduce single subsections for the following sub-results to reflect our main results in a logical order:
  \begin{itemize}
    \item Boltzmann sampling of positively designed sequences for multiple target structures works effectively and accurately, where the full-featured RNA energy model (Turner) can be well approximated.
      For this approximation, we use the much more efficiently tractable stacking model; in this model, controlled Boltzmann sampling for typical multi-target design instances requires little time (corresponding to small tree-width);
    \item Multi-dimensional Boltzmann sampling solves the positive design problem effectively, requiring only a limited number of rejections (even for multiple simultaneously considered features);
    \item Our method can be beneficially used in the setting of negative design. Here, we demonstrate a typical use case of the method. Moreover, we empirically show that 
      the positively designed sequences can improve negative design approaches, where
      they work better than previous approaches (for seed generation).
\end{itemize}}

@@

We included further data comparing to RNABluePrint in the supplement (Supp. Mat. C) and discuss the comparison to MODENA. While comparing to MODENA in the same way as to RNABluePrint can admittedly appear obvious, the sampling in MODENA is of a very different nature, such that---in our opinion---elaborate comparisons would not provide relevant insights. Of course, this has to be explained:

@@

\Answer{
As this reviewer could probably notice, the techniques used in this paper are quite diverse, and stem from several subfields of computer science (combinatorial optimization, graph algorithms, enumerative/analytic combinatorics...). For this reason, a more concise description of the method would be poorly self-contained, possibly to the point of irreproducibility. However, we share the concern of this reviewer regarding an imbalanced manuscript, and moved less original/central parts of the method to the appendix. This includes the expression of various energy models as constraints, and the description of our naive numerical weight optimization.
}
%\AnswerTODO[Yann]{
%Move material to appendix
%Better motivations for \#P-hardness
%New results : 
%  1) Validation of stacking approximation: Correlation coefficient of Stacking and Turner (+ %goodness of fit)
%  2) Tree width and Time benchmark
%  3) Rejection analysis for multidim sampling
%  4) Proof-of-concept -Potential of better seeds
%}
\Comment{
2) Although, it was shown previously, that RNABluePrint is a leading tool for multiple design, it would be valuable to include and compare the results generated by the other tools, e.g. Frnakenstein.}%

@@

\Answer{We were unsure about what was meant in this comment. Just to be precise, we distinguish the term 'Boltzmann sampling' from 'multi-dimensional Bolzmann sampling' in our paper. 
On the one hand, Boltzmann sampling, i.e. drawing sequences from the Boltzmann distribution does not involve any rejection, and is actually exactly as costly as drawing them uniformly. In such a uniform mode, RNARedPrint performs equally as fast as RNABluePrint, and usually faster, since the graph decomposition used in RNABluePrint implies time and memory consumptions that are provably strictly less efficient than the tree decomposition underlying RNARedPrint.
On the other hand, multi-dimensional Boltzmann sampling requires a rejection step to target specific properties that are typically difficult to obtain by chance in the uniform distribution. There is, of course, an added cost associated to this rejection, and the simultaneous weight optimization, which we discuss further in a new "Result" subsection, where we also report the exact time differences.}

>>>>

In particular, the reviews helped us to state the goal of this work more clearly and improve its presentation. While we demonstrate the principle \emph{applicability} of our approach to \emph{negative} RNA design, which  aims at high and similar probability of the target structures (here measured as 'MultiDefect'), the \emph{focus} of this work is the \emph{positive} design of sequences that fold into multiple target structures with highly specific properties. We discuss this positive design, in particular, for specific folding energies (for each single target structure) and GC content. Moreover, as we like to emphasize, the method is even more general, which we see as a very important strength of our approach; it is prepared to handle many additional constraints and sequence features, like they often are required in practical use cases, using the same mechanism.

Naturally, we hope that the method turns out to be highly useful for single- and multi-target negative design (and various other applications, like generating background models for assessing significance) in future applications, but---while we present preliminary results for negative design (based on generating seed sequences for the MODENA benchmark)---a more in-depth evaluation of its use in negative design is beyond the scope of this paper. This seems even more reasonable, since the principal benefits of controlled sampling  have been demonstrated by RNABlueprint. We apologize for potential misunderstandings about the objectives of the manuscript and hope to better clarify these aspects in our revision.

@@

\Answer{
  We thank the reviewer for voicing her concerns, which motivated us to work on clarifying the practical use of the method/tool RNARedPrint in several ways; we have carefully edited the entire manuscript having these issues in mind.
  In the methods section, we tried to improve the general readability especially of our overview of the method and central problem.
  Moreover, we structured and rephrased the Results section more clearly and adapted our contributions outline and conclusion accordingly. In Results, we introduce single subsections for the following sub-results to reflect our main results in a logical order:
  \begin{itemize}
    \item Boltzmann sampling of positively designed sequences for multiple target structures works effectively and accurately, where the full-featured RNA energy model (Turner) can be well approximated.
      For this approximation, we use the much better tractable stacking energy model; in this model, controlled Boltzmann sampling for typical multi-target design instances requires little time (corresponding to small treewidth);
    \item Multi-dimensional Boltzmann sampling solves the positive design problem effectively, requiring only a limited number of iterations (and therefore rejections)---even for multiple simultaneously considered features;
    \item Our method can be beneficially used in the setting of negative design. Here, we demonstrate a typical use case of the method. By detailed analysis of this procedure, we demonstrate the feasibility of the approach for practically relevant design instances. Moreover, we empirically show that the positively designed sequences are suitable to improve negative design approaches, where they yield better performance than seeds generated by previous approaches.
\end{itemize}}

@@

We included further data comparing to RNABluePrint in the supplement (Supp. Mat. C) and discuss the comparison to MODENA. While comparing to MODENA in the same way as to RNABluePrint can admittedly appear obvious, the sampling in MODENA is of a very different nature, such that---in our opinion---elaborate comparisons would not provide relevant insights. Of course, this has to be explained in more detail:

@@

\Answer{
We share the concern of this reviewer that an unncecessarily imbalanced manuscript should be avoided, and moved less original/central parts of the method to the appendix. This includes the expression of various energy models as constraints, and the description of our naive numerical weight optimization. However, we want to emphasize that the techniques used in this paper are quite diverse, and stem from several subfields of computer science (combinatorial optimization, graph algorithms, enumerative/analytic combinatorics...). For this reason, we feel that the method part cannot be substantially shortened beyond the current point without making the text less accessible and poorly self-contained. 
}
%\AnswerTODO[Yann]{
%Move material to appendix
%Better motivations for \#P-hardness
%New results : 
%  1) Validation of stacking approximation: Correlation coefficient of Stacking and Turner (+ %goodness of fit)
%  2) Tree width and Time benchmark
%  3) Rejection analysis for multidim sampling
%  4) Proof-of-concept -Potential of better seeds
%}
\Comment{
2) Although, it was shown previously, that RNABluePrint is a leading tool for multiple design, it would be valuable to include and compare the results generated by the other tools, e.g. Frnakenstein.}%

@@

\Answer{We were not completely sure what was meant by this comment. Just to be precise, we distinguish the term 'Boltzmann sampling' from 'multi-dimensional Bolzmann sampling' in our paper. 
On the one hand, Boltzmann sampling, i.e. drawing sequences from the Boltzmann distribution does not involve any rejection, and is thus exactly as fast as drawing them uniformly. For uniform sampling, RNARedPrint is usually even faster than RNABluePrint, due to the state-of-the-art tree decomposition underlying RNARedPrint, which typically yields lower treewidths than the graph decomposition of RNABluePrint.
On the other hand, multi-dimensional Boltzmann sampling requires a rejection step to target specific properties that are typically difficult to obtain by chance in the uniform distribution. There is, of course, an added cost associated to this rejection, and the simultaneous weight optimization. We do now provide detailed overview of the seed generation procedure, which like our multi-dimensional Boltzmann sampling strategy---runs through several iterations of performing the elementary Boltzmann sampling in Suppl.\ Mat. E. The corresponding Figure (Supp. Fig. 9) shows that typically only few iterations suffice to produce the sequences. In this way, the time for multi-dimensional Boltzmann sampling typically is a small multiple of the time required for uniform sampling.}

### 2019-01-25 RNARedPrint.tex
figure description includes arrows now
sync

\begin{abstractbox}
\begin{abstract}
  \parttitle{Background}
  The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology. Synthetic design approaches profit strongly from effective in-silico methods, \revised{which substantially reduce the need for costly wet-lab experiments.} \\
%
  \parttitle{Results} We devise a novel approach to a central ingredient of most in-silico
  design methods: the generation of sequences \revised{that fold well into multiple target structures.} 
  % 
  \revised{Based on constraint networks, our approach \ourprog{}} supports generic
  Boltzmann-weighted sampling, which enables the positive design of RNA sequences
  with specific free energies (for each of multiple, possibly pseudoknotted, target structures) and \GCb-content.
%
  Moreover, we study general properties of our approach empirically and generate
  biologically relevant multi-target Boltzmann-weighted designs for a
  \revised{established} design benchmark. \revised{Our results demonstrate the effectivity and feasibility of the method in practice
  as well as the benefits of Boltzmann sampling over the previously best multi-target sampling strategy---even for the case of negative design of multi-stable RNAs.}\\
%
  \revised{Besides empirically studies, we finally justify the algorithmic details due to a fundamental theoretic result about multi-stable RNA design, namely the \#P-hardness of the counting of designs.}
%
  \parttitle{Conclusion} \ourprog{} introduces a novel, flexible and---as we demonstrate---effective approach to multi-target RNA design, which promises broad applicability and extensibility.

@@

\begin{figure*}[t]
\begin{center}
    \includegraphics[width=.8\textwidth]{Figs/Workflow}
\end{center}
\caption{General outline of \ourprog{}. From a set of target secondary structures (i), base pairs are merged (ii) into a (base pair) dependency graph (iii) and transformed into a tree decomposition (iv). The tree is then used to compute the partition function, followed by a Boltzmann sampling of valid sequences (v). An adaptive scheme learns weights to achieve targeted energies and \GCb\%, leading to the production of suitable designs (vi). \revised{Note that for simplicity, we assume in this figure that only dependencies between the ends of base pairs are considered to evaluate the energies of structures. Our computations based on a more complex energy model, which considers energy contributions of base pair stacks, require additional dependencies.}
  }
\label{fig:workflow}
\end{figure*}

@@

\revised{First, we show how seed sequences can be generated in a Boltzmann distribution, leading to designs that are substantially more stable that those generated uniformly. As can be seen in Figure~\ref{fig:BoltzmannBenefits} and Supp. Mat. \ref{appsec:immediate-benefits-for-negative-design}, sequences generated in the Boltzmann distribution not only reach lower free-energies than those generated in a uniform setting, but also achieve better Boltzmann probabilities. While the former is expected since the Boltzmann distribution explicitly favors low-energy candidates, the latter is somewhat surprising, since the Boltzmann probability of a target structure could, in principle, decrease under Boltzmann sampling due to the partition function growing faster than the Boltzmann factor. The empirical superiority of Boltzmann sampling appears robust, and is at the foundation of other approaches~\cite{Reinharz2013}.

@@

However, while in principle feasible, sampling in a Boltzmann distribution directly using the Turner energy model may induce extreme computational demands, with tree-widths scaling at least as large as the number of nucleotides in the largest loop. Fortunately, we found that intricacy of the Turner energy model can be circumvented with minimal loss of precision by using a simpler stacking energy model. As shown in Figure~\ref{fig:correlation}, a simple stacking energy model, further described in Supp. Mat.~\ref{appsec:modelparameters}, can be used to well approximate the Turner energy model (correlation coefficient $R=0.99$) for our purpose. This greatly reduces the treewidth, and thus the computational requirements of the whole method even for complex instances.
}

@@

\revised{Based on a general framework and efficient algorithms, we introduced a novel approach to design RNA
sequences while targeting very specific complex properties}. In particular, we describe the targeting of the free energies of multiple target structures and the \GCb-content.
Our method combines an FPT stochastic sampling algorithm
with multi-dimensional Boltzmann sampling over distributions
controlled by expressive RNA energy models. \revised{We demonstrate that the approach, despite of its theoretical hardness, performs well on typical multi-stable RNA design instances in practice, which is a direct consequence of the approximability of Turner energies as well as the systematic algorithmic framework. By conducting a proof-of-concept study 
  on an established benchmark set for negative multi-target RNA design (including pseudoknotted instances), we showcase a typical application of our tool \ourprog{}. In this study, 
  the approach generated significantly better seed sequences than
the previously best available method (uniform sampling). These results strongly suggest that the presented technique for positive design can be highly beneficial in future negative design approaches.
}
%
To substantiate our work additionally, we establish the $\#${\sf P}-hardness of uniform sampling which,
from a complexity-theoretic point of view, motivates the FPT, tree decomposition-based nature of our method.

The presented method enables new possibilities for the sequence generation
in the field of RNA sequence design, allowing to enforce additional
constraints, like \GCb\%, while controlling the energy of
multiple target structures. Thus, it presents a major advance over
previously applied ad-hoc sampling and even efficient uniform sampling
strategies. We demonstrated the practical feasibility of such controlled sequence
generation, including the case of pseudoknotted structures, and studied its use for multi-target RNA design.

%perspectives:
Our framework is extensible to include various more complex sequence constraints, including mandatory/forbidden motifs at specific
positions or anywhere in the designed sequences, by adapting formal language constructs
of Zhou \emph{et al}~\citet{Zhou2013}. In future work, negative design principles could 
be explicitly supported at the generation stage, for instance by
penalizing a set of alternative helices/structures.

One could also envision using positive design as a null model, for instance in order to assess the significance of observed properties. Indeed, the complex set of constraints required in order to achieve a desired function must be equally enforced in RNA design, and to produce random sequences that are known to be constrained at a phenotypic level. \revised{Therefore, one could} use constrained design to assess whether P-values, and \revised{thus} our perception of the statistical significance, are affected by the overly simplistic simple null models (\emph{e.g.} dinucleotide shuffling) used to model random RNAs~\cite{Rivas2017}.

@@

\section*{Acknowledgements}
YP is supported
%jointly supported
by the
% French
{\em Agence
  Nationale de la Recherche} and the
%Austrian
FWF (ANR-14-CE34-0011; project RNALands).  SH is supported by the
German Federal Ministry of Education and Research (BMBF support code
031A538B; de.NBI: German Network for Bioinformatics Infrastructure)
and the
% acknowledged by
%financial support of the
Future and Emerging Technologies programme
% within the Seventh Framework Programme for Research of the
%European Commission, under
(FET-Open grant 323987; project RiboNets).
%
We thank Leonid Chindelevitch for suggesting a drastic optimization of our FPT algorithm based on stable sets combinatorics, and Arie Koster for practical recommendations on tree decompositions.

>>>>

\begin{abstractbox}
\begin{abstract}
  \parttitle{Background}
  The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology. Synthetic design approaches profit strongly from effective in-silico methods, \revised{which substantially reduce the need for costly wet-lab experiments.} \\
%
  \parttitle{Results} We devise a novel approach to a central ingredient of most in-silico
  design methods: the generation of sequences \revised{that fold well into multiple target structures.} 
  % 
  \revised{Based on constraint networks, our approach \ourprog{}} supports generic
  Boltzmann-weighted sampling, which enables the positive design of RNA sequences
  with specific free energies (for each of multiple, possibly pseudoknotted, target structures) and \GCb-content.
%
  Moreover, we study general properties of our approach empirically and generate
  biologically relevant multi-target Boltzmann-weighted designs for a
  \revised{established} design benchmark. \revised{Our results demonstrate the effectivity and feasibility of the method in practice
  as well as the benefits of Boltzmann sampling over the previously best multi-target sampling strategy---even for the case of negative design of multi-stable RNAs.}
%
  \revised{Besides empirically studies, we finally justify the algorithmic details due to a fundamental theoretic result about multi-stable RNA design, namely the \#P-hardness of the counting of designs.}
%
  \parttitle{Conclusion} \ourprog{} introduces a novel, flexible and---as we demonstrate---effective approach to multi-target RNA design, which promises broad applicability and extensibility.

@@

\begin{figure*}[t]
\begin{center}
    \includegraphics[width=.8\textwidth]{Figs/Workflow}
\end{center}
\caption{General outline of \ourprog{}. From a set of target secondary structures (i), base pairs are merged (ii) into a (base pair) dependency graph (iii) and transformed into a tree decomposition (iv). The tree is then used to compute the partition function, followed by a Boltzmann sampling of valid sequences (v). An adaptive scheme learns weights to achieve targeted energies and \GCb\% \revised{(arrows)}, leading to the production of suitable designs (vi). \revised{Note that for simplicity, we assume in this figure that only dependencies between the ends of base pairs are considered to evaluate the energies of structures. Our computations based on a more complex energy model, which considers energy contributions of base pair stacks, require additional dependencies.}
  }
\label{fig:workflow}
\end{figure*}

@@

\revised{First, we show how seed sequences can be generated in a Boltzmann distribution, leading to designs that are substantially more stable that those generated uniformly. As can be seen in Figure~\ref{fig:BoltzmannBenefits} and Supp. Mat. \ref{appsec:immediate-benefits-for-negative-design}, sequences generated in the Boltzmann distribution not only reach lower free-energies than those generated in a uniform setting, but also achieve better Boltzmann probabilities. While the former is expected since the Boltzmann distribution explicitly favors low-energy candidates, the latter is somewhat surprising, since the Boltzmann probability of a target structure could, in principle, decrease under Boltzmann sampling due to the partition function growing faster than the Boltzmann factor. The empirical superiority of Boltzmann sampling appears robust to the target structure length and topology, as demonstrated by prior work~\cite{Reinharz2013}.

@@

However, while in principle feasible, sampling in a Boltzmann distribution directly using the Turner energy model may induce extreme computational demands, with treewidths scaling at least as large as the number of nucleotides in the largest loop. Fortunately, we found that intricacy of the Turner energy model can be circumvented with minimal loss of precision by using a simpler stacking energy model. As shown in Figure~\ref{fig:correlation}, a simple stacking energy model, whose design principles are further described in Supp. Mat.~\ref{appsec:modelparameters}, can be used to approximate the Turner energy model very adequately (correlation coefficient $R=0.99$) in the context of sequence design. Using this simpler model greatly reduces the treewidth, and thus the computational requirements of the whole method even for complex instances.
}

@@

\revised{Based on a general framework and efficient algorithms, we introduce a novel approach to design RNA
sequences while targeting very specific complex properties}. In particular, we describe the targeting of the free energies of multiple target structures and the \GCb-content.
%
Our method combines a fixed-parameter tractable (FPT) sampling algorithm
with multi-dimensional Boltzmann sampling over distributions
controlled by expressive RNA energy models. \revised{We demonstrate that the approach, despite of its theoretical hardness, performs well on typical multi-stable RNA design instances in practice. This good performance is a direct consequence of the approximability of Turner energies as well as the systematic algorithmic framework. By conducting a proof-of-concept study 
  on an established benchmark set for negative multi-target RNA design (including pseudoknotted instances), we showcase a typical application of our tool \ourprog{}. In this study, 
  the approach generates significantly better seed sequences than
the previously best available method (uniform sampling). Our results strongly suggest that the presented technique for positive design can be highly beneficial in future negative design approaches.
}%
%
% SW--we have written most of this already:
%In this way, we as well demonstrated the practical feasibility of such controlled %sequence generation, including the case of pseudoknotted structures.
% YP--Stick to the present tense if you write anything new :) (or move everything to past) 
%
%
To substantiate our work additionally, we establish the $\#${\sf P}-hardness of uniform sampling which,
from a complexity-theoretic point of view, motivates the FPT, tree decomposition-based nature of our method.

\revised{In this way, our framework enables new possibilities
in the field of RNA sequence design.}
%allowing to enforce additional
%constraints, like \GCb\%, while controlling the energy of
%multiple target structures.
%Thus, it presents a major advance over
%previously applied ad-hoc sampling and even efficient %uniform sampling
%strategies. 
%
%perspectives:
\revised{As particular advantage over previous sequence generation methods,} it is extensible to include various more complex sequence constraints, including mandatory/forbidden motifs at specific
positions or anywhere in the designed sequences, by adapting formal language constructs
of Zhou \emph{et al}~\citet{Zhou2013}. In future work, negative design principles could 
be explicitly supported at the generation stage, for instance by
penalizing a set of alternative helices/structures.
%
\revised{We moreover envision using positive design to assess the significance of observed properties. Critically, our current perception of statisitical significance suffers from overly simplistic simple null models (\emph{e.g.} dinucleotide shuffling) used to model random RNAs~\cite{Rivas2017}). Here, our approach promises fundamental improvements by constructing null models of random sequences that satisfy multiple complex constraints.}

@@

\section*{Acknowledgements}
YP is supported
%jointly supported
by the
% French
{\em Agence
  Nationale de la Recherche} and the
%Austrian
Austrian Science Fund (ANR-14-CE34-0011 and FWF-I-1804-N28; project RNALands).  SH is supported by the
German Federal Ministry of Education and Research (BMBF support code
031A538B; de.NBI: German Network for Bioinformatics Infrastructure)
and the
% acknowledged by
%financial support of the
Future and Emerging Technologies programme
% within the Seventh Framework Programme for Research of the
%European Commission, under
(FET-Open grant 323987; project RiboNets).
%
We thank Leonid Chindelevitch for suggesting a drastic optimization of our FPT algorithm based on stable sets combinatorics, and Arie Koster for practical recommendations on tree decompositions.
